Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-08-08 paper/paper.tex
Unboxing with update of Section 3.3

\subsection{Putting everything together}
\subsubsection{Consequences for the DMN function}
The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.
The more familiar and predictable the current environment, the more brain resources may remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context.
%
This speculation receives quantitative support in that connectional links between
nodes of the DMN have been reported to be more consistent and reliable than
functional couplings within any other macroscopical networks \citep{shehz2009}.
As such, random-sampling-related baseline evaluation of possibilities and
events may be subserved by the DMN and get partly suspended when novelty in
the external environment is encountered or immediate action is required.
%
In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making has led the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing intervention of the organism on the world in general. Among all parts of the DMN, the RTPJ is perhaps the most evident candidate for a network-switching relay that calibrates between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

Additionally,
the DMN was proposed to be situated
at the top of the brain network hierarchy, with
the subordinate salience and dorsal attention network in the middle and
the primary sensory cortices at the bottom
\citep{carhart2010default, margulies2016situating}.
Its putative involvement in thinking about 
hypothetical experiences and future outcomes
appears to tie in with the implicit computation of
action and state cascades as a function of
experienced events and collected feedback from the past.
A policy matrix encapsulates the choice probabilities of possible actions
on the world given a current situation (i.e., state).
The DMN may subserve
  constant exploration of candidate action trajectories and
  nested estimaton of their
  cumulative reward outcomes. Implicit computation of future choices
  provides a potential explanation for the
  evolutionary emergence and practical usefulness of
  mind-wandering at day-time and dreams during sleep
  in humans.
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{\textbf{Default mode network:
  possible neurobiological implementation of reinforcement learning.}
  Overview of how the constituent regions of the DMN (refer to section \ref{sec:nodes}; blue horizontal dashed line indicates the border between BA9 and BA10) may
  map onto computational components necessary for an RL agent.
  %
  Axonal tracing in monkeys and diffusion tractography in humans suggested
  that the NAc of the reward circuitry has monosynaptic fiber connections
  to the vmPFC \citep{haber1995, croxson2005quantitative}. Evaluation of
  propagated value information and triggered affective states encoded
  in the vmPFC may then feed into the functionally connected partner nodes of
  the DMN, such as the dmPFC and PMC \citep{andrews2010, bzdok2013segregation}.
}
  \label{fig:rl_process_chart}
\end{figure}

@@

@@

@@

\subsubsection{Experimental predictions}
Following our process model of DMN, several empirical predictions can be drawn.

\textbf{Neural correlates of the Bellman equation in the DMN (Humans)}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscientific insights,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.

\textbf{Continuous learning paradigm (Humans)} We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment.

\textbf{Temporal discounting experiment (Humans)} We hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced-back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

\textbf{Synaptic epigenesis and creation of new entry in policy matrix (Humans \& Animals)} An additional layer of learning concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synapti\c changes compare to other parts of the brain.

\textbf{Interplay with memory and “mind-searching” (Animals)} We hypothesize that experience replay for browsing problem solutions subserved by the DMN is contributing to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. Necessity of cortical DMN regions, in addition to the hippocampus, for “mind-searching” candidate actions during choice behavior can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

% \end{enumerate}
% \end{mdframed}

>>>>

\subsection{Putting everything together}

The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.
The more familiar and predictable the current environment, the more brain resources may remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context.
%
This speculation receives quantitative support in that connectional links between
nodes of the DMN have been reported to be more consistent and reliable than
functional couplings within any other macroscopical networks \citep{shehz2009}.
As such, random-sampling-related baseline evaluation of possibilities and
events may be subserved by the DMN and get partly suspended when novelty in
the external environment is encountered or immediate action is required.
%
In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making has led the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing intervention of the organism on the world in general. Among all parts of the DMN, the RTPJ is perhaps the most evident candidate for a network-switching relay that calibrates between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

Additionally,
the DMN was proposed to be situated
at the top of the brain network hierarchy, with
the subordinate salience and dorsal attention network in the middle and
the primary sensory cortices at the bottom
\citep{carhart2010default, margulies2016situating}.
Its putative involvement in thinking about 
hypothetical experiences and future outcomes
appears to tie in with the implicit computation of
action and state cascades as a function of
experienced events and collected feedback from the past.
A policy matrix encapsulates the choice probabilities of possible actions
on the world given a current situation (i.e., state).
The DMN may subserve
  constant exploration of candidate action trajectories and
  nested estimaton of their
  cumulative reward outcomes. Implicit computation of future choices
  provides a potential explanation for the
  evolutionary emergence and practical usefulness of
  mind-wandering at day-time and dreams during sleep
  in humans.

\paragraph*{Neural correlates of the Bellman equation in the DMN (Humans)}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscientific insights,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.

\paragraph*{Continuous learning paradigm (Humans)} We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment.

\paragraph*{Temporal discounting experiment (Humans)} We hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced-back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{\textbf{Default mode network:
  possible neurobiological implementation of reinforcement learning.}
  Overview of how the constituent regions of the DMN (refer to section \ref{sec:nodes}; blue horizontal dashed line indicates the border between BA9 and BA10) may
  map onto computational components necessary for an RL agent.
  %
  Axonal tracing in monkeys and diffusion tractography in humans suggested
  that the NAc of the reward circuitry has monosynaptic fiber connections
  to the vmPFC \citep{haber1995, croxson2005quantitative}. Evaluation of
  propagated value information and triggered affective states encoded
  in the vmPFC may then feed into the functionally connected partner nodes of
  the DMN, such as the dmPFC and PMC \citep{andrews2010, bzdok2013segregation}.
}
  \label{fig:rl_process_chart}
\end{figure}

@@

\paragraph*{Interplay with memory and “mind-searching” (Animals)} We hypothesize that experience replay for browsing problem solutions subserved by the DMN is contributing to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. Necessity of cortical DMN regions, in addition to the hippocampus, for “mind-searching” candidate actions during choice behavior can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

@@

\paragraph*{Synaptic epigenesis and creation of new entry in policy matrix (Humans \& Animals)} An additional layer of learning concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synapti\c changes compare to other parts of the brain.

@@
